---
title: "Prokaryotic Metatranscriptomic Analysis of Riverine and Phosphorus Additions"
author: "Erick Delgadillo"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    number_sections: true
---

# Overview

This document reproduces the prokaryotic metatranscriptomic analysis
used in the published study on microbial responses to riverine and
phosphorus additions.

The analysis focuses on the non-ribosomal RNA (NoRibo) fraction of the
control (C), river-water addition (R), and river-water plus phosphorus
addition (R+P) mesocosms at 0 h and 72 h after treatment.

The workflow includes taxonomic and functional annotation, transcript
abundance processing, TPM-based analyses, and differential gene
expression analysis.

The original laboratory scripts contained data and analyses from two
independent experiments. This reproducibility workflow retains only the
samples and analyses corresponding to the experiment reported in the
associated publication.

```{r setup, include=FALSE}
# -------------------------------------------------------------------------
# Packages and analysis settings
# -------------------------------------------------------------------------

library(dplyr, warn.conflicts = FALSE)
library(ggplot2)

source(if (file.exists("../R/paths.R")) "../R/paths.R" else "R/paths.R")

SIGNIFICANCE <- 0.05
N_REPLICATES <- 3

# processed_dir <- here::here("data", "processed", "prokaryotes")
# results_dir <- here::here("results", "prokaryotes")
# 
# dir.create(processed_dir, recursive = TRUE, showWarnings = FALSE)
# dir.create(results_dir, recursive = TRUE, showWarnings = FALSE)
```

```{r environmental-data}
# -------------------------------------------------------------------------
# Environmental metadata
# -------------------------------------------------------------------------
# Bacterial production, primary production and nutrient concentrations are
# summarized for the treatment × time combinations used in the published
# experiment.

production_data <- read.csv(
    data_file("metadata/environmental/WP2_Spring_BP_PP_Mesocosmos.csv"),
  sep = ",",
  check.names = FALSE
) %>%
  tidyr::separate(
    sample,
    into = c("treatment", "time"),
    sep = "_"
  ) %>%
  dplyr::select(-errorstandar)

nutrients <- read.csv(
    data_file("metadata/environmental/WP2_Spring_Mesocosmos.csv"),
  sep = ";",
  check.names = FALSE
) %>%
  tidyr::separate(
    sample,
    into = c("treatment", "time"),
    sep = "_"
  ) %>%
  dplyr::filter(time %in% c("0h", "72h")) %>%
  dplyr::group_by(treatment, time, variable) %>%
  dplyr::summarise(
    mean = mean(value),
    .groups = "drop"
  )

Environmental_data <- dplyr::bind_rows(
  nutrients,
  production_data
) %>%
  dplyr::mutate(
    treatment = dplyr::recode(treatment, "RP" = "R+P")
  ) %>%
  tidyr::unite(
    sample,
    treatment,
    time,
    sep = "_",
    remove = TRUE
  ) %>%
  tidyr::pivot_wider(
    names_from = variable,
    values_from = mean
  )
```

# Taxonomic annotation

```{r taxonomy-load}
# -------------------------------------------------------------------------
# Load and clean GTDB taxonomic assignments
# -------------------------------------------------------------------------
# Placeholder and explicitly unclassified assignments are treated as missing,
# allowing downstream grouping to fall back to broader taxonomic ranks.

taxonomy <- readr::read_tsv(
    data_file("raw/prokaryotes/taxonomy_classification_gtdb.tsv"),
  show_col_types = FALSE
) %>%
  dplyr::select(
    orf,
    Domain,
    Phylum,
    Class,
    Order,
    Family,
    Genus,
    Species
  ) %>%
  dplyr::mutate(
    dplyr::across(
      c(Domain, Phylum, Class, Order, Family, Genus, Species),
      ~ ifelse(
          stringr::str_detect(
            .x,
            "_X|_[A-I]|uncl\\.|Unclassified"
          ),
          NA,
          .x
        )
    )
  )

orders_of_interest <- c(
  "Rhodobacterales",
  "Puniceispirillales",
  "Rhizobiales",
  "Pelagibacterales",
  "Pseudomonadales",
  "Enterobacterales",
  "Burkholderiales",
  "Beggiatoales",
  "SAR86",
  "Flavobacteriales"
)

classes_of_interest <- c(
  "Alphaproteobacteria",
  "Gammaproteobacteria"
)
```

```{r taxonomy-consensus}
# -------------------------------------------------------------------------
# Hierarchical consensus for duplicated ORF assignments
# -------------------------------------------------------------------------
# Some ORFs have more than one taxonomic assignment. Missing information in
# one assignment is not considered a conflict. A conflict occurs only when
# different non-missing values are assigned to the same rank.

Taxonomy_consensus <- taxonomy %>%
  dplyr::group_by(orf) %>%
  dplyr::summarise(
    conflict_Domain = dplyr::n_distinct(Domain, na.rm = TRUE) > 1,
    Domain = dplyr::first(
      Domain[!is.na(Domain)],
      default = NA_character_
    ),

    conflict_Phylum = dplyr::n_distinct(Phylum, na.rm = TRUE) > 1,
    Phylum = dplyr::first(
      Phylum[!is.na(Phylum)],
      default = NA_character_
    ),

    conflict_Class = dplyr::n_distinct(Class, na.rm = TRUE) > 1,
    Class = dplyr::first(
      Class[!is.na(Class)],
      default = NA_character_
    ),

    conflict_Order = dplyr::n_distinct(Order, na.rm = TRUE) > 1,
    Order = dplyr::first(
      Order[!is.na(Order)],
      default = NA_character_
    ),

    conflict_Family = dplyr::n_distinct(Family, na.rm = TRUE) > 1,
    Family = dplyr::first(
      Family[!is.na(Family)],
      default = NA_character_
    ),

    conflict_Genus = dplyr::n_distinct(Genus, na.rm = TRUE) > 1,
    Genus = dplyr::first(
      Genus[!is.na(Genus)],
      default = NA_character_
    ),

    conflict_Species = dplyr::n_distinct(Species, na.rm = TRUE) > 1,
    Species = dplyr::first(
      Species[!is.na(Species)],
      default = NA_character_
    ),

    .groups = "drop"
  ) %>%
  dplyr::mutate(
    Domain = dplyr::if_else(
      conflict_Domain,
      NA_character_,
      Domain
    ),

    Phylum = dplyr::if_else(
      conflict_Domain | conflict_Phylum,
      NA_character_,
      Phylum
    ),

    Class = dplyr::if_else(
      conflict_Domain | conflict_Phylum | conflict_Class,
      NA_character_,
      Class
    ),

    Order = dplyr::if_else(
      conflict_Domain | conflict_Phylum | conflict_Class |
        conflict_Order,
      NA_character_,
      Order
    ),

    Family = dplyr::if_else(
      conflict_Domain | conflict_Phylum | conflict_Class |
        conflict_Order | conflict_Family,
      NA_character_,
      Family
    ),

    Genus = dplyr::if_else(
      conflict_Domain | conflict_Phylum | conflict_Class |
        conflict_Order | conflict_Family | conflict_Genus,
      NA_character_,
      Genus
    ),

    Species = dplyr::if_else(
      conflict_Domain | conflict_Phylum | conflict_Class |
        conflict_Order | conflict_Family | conflict_Genus |
        conflict_Species,
      NA_character_,
      Species
    )
  )
```

```{r taxonomy-final}
# -------------------------------------------------------------------------
# Final taxonomic grouping
# -------------------------------------------------------------------------
# Taxonomic resolution is intentionally heterogeneous. Ecologically relevant
# groups are retained at order level, while other organisms are summarized at
# class, phylum or domain level.

Taxonomy <- Taxonomy_consensus %>%
  dplyr::mutate(
    Final_Taxonomy = dplyr::case_when(
      Domain == "Archaea" ~ Domain,

      Phylum == "Actinobacteriota" ~ Phylum,
      Phylum == "Spirochaetes" ~ Phylum,
      Phylum == "Firmicutes" ~ Phylum,
      Phylum == "Cyanobacteria" ~ Phylum,
      Phylum == "Planctomycetota" ~ Phylum,

      Class == "Alphaproteobacteria" &
        Order %in% orders_of_interest ~ Order,

      Class == "Alphaproteobacteria" &
        !Order %in% orders_of_interest ~ Class,

      Class == "Gammaproteobacteria" &
        Order %in% orders_of_interest ~ Order,

      Class == "Gammaproteobacteria" &
        !Order %in% orders_of_interest ~ Class,

      Phylum == "Bacteroidota" &
        Order %in% orders_of_interest ~ Order,

      Phylum == "Bacteroidota" &
        !Order %in% orders_of_interest ~ Phylum,

      Phylum == "Proteobacteria" &
        !Class %in% classes_of_interest ~ Phylum,

      TRUE ~ Domain
    )
  ) %>%
  dplyr::select(
    orf,
    Domain,
    Phylum,
    Class,
    Order,
    Family,
    Genus,
    Species,
    Final_Taxonomy
  )

stopifnot(!anyDuplicated(Taxonomy$orf))
```

# Sample metadata

```{r sample-metadata}
# -------------------------------------------------------------------------
# Prokaryotic sample metadata
# -------------------------------------------------------------------------
# Sequencing identifiers are linked to the experimental design. Only samples
# belonging to the experiment reproduced in this repository are retained.

samples_tags <- read.csv(
  data_file("metadata/prokaryotes/INTERES_Prok_samples_tags.csv"),
  sep = ",",
  check.names = FALSE
) %>%
  dplyr::select(
    sample,
    Sample,
    T1
  ) %>%
  tidyr::unite(
    sample,
    sample,
    T1,
    sep = "_",
    remove = TRUE
  ) %>%
  tidyr::separate(
    Sample,
    into = c(
      "Workpackage",
      "experiment",
      "treatment",
      "time",
      "replic"
    ),
    sep = "_",
    remove = FALSE
  ) %>%
  dplyr::filter(Workpackage == "WP2")

stopifnot(
  nrow(samples_tags) == 18,
  !anyDuplicated(samples_tags$sample)
)
```

# Functional annotation

```{r functional-annotation}
# -------------------------------------------------------------------------
# eggNOG functional annotation
# -------------------------------------------------------------------------
# Only the identifiers required downstream are retained. Up to three KEGG
# ortholog assignments are preserved in priority order.

Prok_annotations <- suppressWarnings(
  data.table::fread(
      data_file("raw/prokaryotes/all_samples_emapper_annotations.tsv"))
) %>%
  dplyr::rename(
    orf = `#query`
  ) %>%
  dplyr::select(
    orf,
    KEGG_ko,
    KEGG_Module,
    PFAMs
  ) %>%
  tidyr::separate(
    KEGG_ko,
    into = c("KEGG_ko", "KEGG_ko2", "KEGG_ko3"),
    sep = ",",
    remove = TRUE,
    extra = "drop",
    fill = "right"
  ) %>%
  dplyr::mutate(
    KEGG_ko = stringr::str_remove(KEGG_ko, "ko:"),
    KEGG_ko2 = stringr::str_remove(KEGG_ko2, "ko:"),
    KEGG_ko3 = stringr::str_remove(KEGG_ko3, "ko:"),
    KEGG_Module = sub(",.*$", "", KEGG_Module)
  )

stopifnot(!anyDuplicated(Prok_annotations$orf))
```

# Transcript abundance

```{r counts-and-tpm}
# -------------------------------------------------------------------------
# Raw counts, prokaryotic filtering and TPM recalculation
# -------------------------------------------------------------------------
# Counts are restricted to the 18 study samples and to ORFs represented in
# the curated prokaryotic taxonomy. TPM is recalculated within this retained
# prokaryotic fraction.

prok_counts_raw <- readr::read_tsv(
    data_file("raw/prokaryotes/counts.tsv.gz"),
  show_col_types = FALSE
) %>%
  dplyr::filter(
    sample %in% samples_tags$sample
  ) %>%
  dplyr::semi_join(
    Taxonomy %>%
      dplyr::filter(Domain != "Eukaryota"),
    by = "orf"
  ) %>%
  dplyr::group_by(sample) %>%
  dplyr::mutate(
    tpm = (count / sum(count)) * 1e6
  ) %>%
  dplyr::ungroup()

stopifnot(
  setequal(unique(prok_counts_raw$sample), samples_tags$sample)
)

tpm_check <- prok_counts_raw %>%
  dplyr::group_by(sample) %>%
  dplyr::summarise(
    total_tpm = sum(tpm),
    .groups = "drop"
  )

stopifnot(
  all(abs(tpm_check$total_tpm - 1e6) < 1)
)
```

# Protein profiles of interest

```{r protein-profiles}
# -------------------------------------------------------------------------
# Functional protein profiles
# -------------------------------------------------------------------------
# Descriptive names not used downstream are discarded before duplicate
# removal. This prevents duplicated TPM values when two profile rows differ
# only in the unused `name` field.

Hiper_proteins_analysis <- readxl::read_excel(
    data_file("metadata/environmental/Protein_profiles_2024.xlsx"),
  sheet = "Protein profiles",
  col_names = TRUE
) %>%
  dplyr::select(
    Category,
    Subcategory,
    SubcategoryP,
    gene_protein,
    KEGG_ko,
    kingdom
  ) %>%
  dplyr::filter(!is.na(KEGG_ko)) %>%
  dplyr::distinct()

functional_kos <- unique(Hiper_proteins_analysis$KEGG_ko)
```

# Integration of expression and annotations

```{r annotation-joins}
# -------------------------------------------------------------------------
# Join expression, taxonomy, functional annotation and sample metadata
# -------------------------------------------------------------------------

Prok_annotated <- prok_counts_raw %>%
  dplyr::left_join(
    Taxonomy,
    by = "orf"
  ) %>%
  dplyr::left_join(
    Prok_annotations,
    by = "orf"
  ) %>%
  dplyr::left_join(
    samples_tags,
    by = "sample"
  )

stopifnot(
  nrow(Prok_annotated) == nrow(prok_counts_raw)
)
```

```{r annotated-counts-tpm}
# -------------------------------------------------------------------------
# Tables used by downstream analyses
# -------------------------------------------------------------------------
# There is one observation per ORF and biological replicate at this stage,
# so the original group_by()/summarise() operations reduce to a direct
# projection of the required columns.

Prok_counts_annotated <- Prok_annotated %>%
  dplyr::transmute(
    Workpackage,
    treatment,
    time,
    replic,
    orf,
    Final_Taxonomy,
    Phylum,
    Class,
    Order,
    Family,
    Genus,
    Species,
    KEGG_ko,
    KEGG_ko2,
    KEGG_ko3,
    KEGG_Module,
    PFAMs,
    count,
    nReplics = 1L
  )

Prok_TPMs_annotated <- Prok_annotated %>%
  dplyr::transmute(
    Workpackage,
    treatment,
    time,
    replic,
    orf,
    Final_Taxonomy,
    Domain,
    Phylum,
    Class,
    Order,
    Family,
    Genus,
    Species,
    KEGG_ko,
    KEGG_ko2,
    KEGG_ko3,
    KEGG_Module,
    PFAMs,
    tpm,
    nReplics = 1L
  )

arrow::write_parquet(
  Prok_counts_annotated,
    data_file("processed/prokaryotes/prok_counts_annotated.parquet"))


arrow::write_parquet(
  Prok_TPMs_annotated,
    data_file("processed/prokaryotes/prok_tpms_annotated.parquet"))
```

# Functional TPM analysis

```{r functional-selection}
# -------------------------------------------------------------------------
# Select genes of functional interest using prioritized KEGG assignments
# -------------------------------------------------------------------------
# For each ORF, the first matching KEGG assignment is retained in the order
# KEGG_ko -> KEGG_ko2 -> KEGG_ko3.

Prok_filtered <- Prok_TPMs_annotated %>%
  dplyr::mutate(
    selected_KEGG = dplyr::case_when(
      KEGG_ko %in% functional_kos ~ KEGG_ko,
      KEGG_ko2 %in% functional_kos ~ KEGG_ko2,
      KEGG_ko3 %in% functional_kos ~ KEGG_ko3,
      TRUE ~ NA_character_
    )
  ) %>%
  dplyr::filter(!is.na(selected_KEGG)) %>%
  dplyr::mutate(
    KEGG_ko = selected_KEGG
  ) %>%
  dplyr::select(-selected_KEGG)

Prok_functional_joined <- Prok_filtered %>%
  dplyr::left_join(
    Hiper_proteins_analysis,
    by = "KEGG_ko",
    relationship = "many-to-many"
  ) %>%
  dplyr::filter(
    kingdom != "Eukarya"
  )
```

```{r functional-tpm}
# -------------------------------------------------------------------------
# Aggregate TPM by functional gene and taxonomic group
# -------------------------------------------------------------------------

Prok_functional_TPM <- Prok_functional_joined %>%
  dplyr::group_by(
    Workpackage,
    treatment,
    time,
    replic,
    gene_protein,
    KEGG_ko,
    Category,
    Subcategory,
    SubcategoryP,
    Final_Taxonomy
  ) %>%
  dplyr::summarise(
    tpm = sum(tpm),
    .groups = "drop"
  )

arrow::write_parquet(
  Prok_functional_TPM,
    data_file("processed/prokaryotes/prok_functional_tpms.parquet"))
```

# Mean TPM and taxonomic composition

```{r mean-tpm}
# -------------------------------------------------------------------------
# Mean TPM across three biological replicates
# -------------------------------------------------------------------------
# ORFs absent from a replicate are implicit zeros in the long table.
# Therefore, observed TPM values are summed and divided by the fixed number
# of biological replicates rather than averaged only across detected rows.

Prok_mean_TPM <- Prok_TPMs_annotated %>%
  dplyr::group_by(
    Workpackage,
    treatment,
    time,
    orf,
    Final_Taxonomy,
    Phylum,
    Class,
    Order,
    Family,
    Genus,
    Species,
    KEGG_ko,
    KEGG_ko2,
    KEGG_ko3,
    KEGG_Module,
    PFAMs
  ) %>%
  dplyr::summarise(
    tpm = sum(tpm) / N_REPLICATES,
    nReplics_detected = dplyr::n(),
    .groups = "drop"
  )

arrow::write_parquet(
  Prok_mean_TPM,
    data_file("processed/prokaryotes/prok_mean_tpms.parquet"))
```

```{r taxonomic-composition}
# -------------------------------------------------------------------------
# Taxonomic composition of the prokaryotic metatranscriptome
# -------------------------------------------------------------------------

Prok_MetaT_Proportion_FinalTax <- Prok_mean_TPM %>%
  tidyr::unite(
    sample,
    treatment,
    time,
    sep = "_",
    remove = FALSE
  ) %>%
  dplyr::group_by(
    sample,
    Final_Taxonomy
  ) %>%
  dplyr::summarise(
    Total_tpm = sum(tpm),
    .groups = "drop"
  )
```

# Functional gene ratios

```{r gene-ratios}
# -------------------------------------------------------------------------
# Functional gene ratios within taxonomic groups
# -------------------------------------------------------------------------
# Functional TPM values are averaged across the three biological replicates.
# This is algebraically equivalent to first averaging each ORF and then
# repeating the KEGG-selection step because functional annotation is fixed
# for each ORF across replicates.
#
# pstA, pstB, pstC and pstS are grouped as components of the pstABCS
# high-affinity phosphate transport system.

Prok_gene_ratios <- Prok_functional_TPM %>%
  dplyr::mutate(
    gene_protein = dplyr::if_else(
      gene_protein %in% c("pstA", "pstB", "pstC", "pstS"),
      "pstABCS",
      gene_protein
    )
  ) %>%
  dplyr::group_by(
    treatment,
    time,
    gene_protein,
    Category,
    Subcategory,
    SubcategoryP,
    Final_Taxonomy
  ) %>%
  dplyr::summarise(
    tpm = sum(tpm) / N_REPLICATES,
    .groups = "drop"
  ) %>%
  tidyr::unite(
    sample,
    treatment,
    time,
    sep = "_",
    remove = TRUE
  ) %>%
  dplyr::left_join(
    Prok_MetaT_Proportion_FinalTax,
    by = c("sample", "Final_Taxonomy")
  ) %>%
  dplyr::mutate(
    GenRatio = tpm / Total_tpm
  ) %>%
  dplyr::left_join(
    Environmental_data,
    by = "sample"
  )

stopifnot(
  !any(is.na(Prok_gene_ratios$Total_tpm)),
  !any(is.na(Prok_gene_ratios$GenRatio)),
  !any(is.infinite(Prok_gene_ratios$GenRatio))
)

readr::write_csv(
  Prok_gene_ratios,
    data_file("processed/prokaryotes/prok_gene_ratios.csv"))
```

# Differential gene expression analysis

Differential gene expression was evaluated from raw prokaryotic read
counts using edgeR. Only samples belonging to the experiment described
in the associated publication are included in this analysis.

```{r dge-count-matrix}
# -------------------------------------------------------------------------
# Count matrix and experimental groups
# -------------------------------------------------------------------------
# ORFs not detected in a sample are assigned a count of zero. Sample columns
# are explicitly ordered according to the metadata.

Prok_count_matrix <- prok_counts_raw %>%
  dplyr::select(
    orf,
    sample,
    count
  ) %>%
  tidyr::pivot_wider(
    names_from = sample,
    values_from = count,
    values_fill = 0
  ) %>%
  tibble::column_to_rownames("orf")

Prok_count_matrix <- Prok_count_matrix[
  ,
  samples_tags$sample,
  drop = FALSE
]

dge_samples <- samples_tags %>%
  dplyr::mutate(
    group = factor(
      paste(treatment, time, sep = "_"),
      levels = c(
        "C_0h",
        "R_0h",
        "R+P_0h",
        "C_72h",
        "R_72h",
        "R+P_72h"
      )
    )
  )

stopifnot(
  identical(
    colnames(Prok_count_matrix),
    dge_samples$sample
  )
)
```

```{r dge-model}
# -------------------------------------------------------------------------
# edgeR filtering, normalization and model fitting
# -------------------------------------------------------------------------

dge_samples_df <- as.data.frame(dge_samples)
rownames(dge_samples_df) <- dge_samples_df$sample

dgelist <- edgeR::DGEList(
  counts = Prok_count_matrix,
  samples = dge_samples_df
)

keep <- edgeR::filterByExpr(
  dgelist,
  group = dgelist$samples$group
)

dgelist <- dgelist[
  keep,
  ,
  keep.lib.sizes = FALSE
]

dgelist <- edgeR::calcNormFactors(
  dgelist,
  method = "TMM"
)

design <- model.matrix(
  ~ 0 + group,
  data = dgelist$samples
)

colnames(design) <- c(
  "C_0h",
  "R_0h",
  "RP_0h",
  "C_72h",
  "R_72h",
  "RP_72h"
)

stopifnot(
  identical(
    rownames(design),
    rownames(dgelist$samples)
  )
)

cdisp <- edgeR::estimateDisp(
  dgelist,
  design
)

fit <- edgeR::glmQLFit(
  cdisp,
  design
)
```

```{r dge-contrasts}
# -------------------------------------------------------------------------
# Differential expression contrasts
# -------------------------------------------------------------------------
# Positive logFC values indicate higher expression in the first group named
# in each contrast.

contrasts <- limma::makeContrasts(
  R_vs_C_0h = R_0h - C_0h,
  RP_vs_C_0h = RP_0h - C_0h,
  RP_vs_R_0h = RP_0h - R_0h,
  R_vs_C_72h = R_72h - C_72h,
  RP_vs_C_72h = RP_72h - C_72h,
  RP_vs_R_72h = RP_72h - R_72h,
  levels = design
)
```

```{r dge-tests}
# -------------------------------------------------------------------------
# Quasi-likelihood tests and FDR correction
# -------------------------------------------------------------------------
# The six explicitly defined contrasts are tested in the order shown above.

Prok_DGE <- tibble::tibble()

for (contrast_name in colnames(contrasts)) {

  test <- edgeR::glmQLFTest(
    fit,
    contrast = contrasts[, contrast_name]
  )

  contrast_result <- edgeR::topTags(
    test,
    n = Inf,
    adjust.method = "fdr"
  )$table %>%
    tibble::rownames_to_column("orf") %>%
    dplyr::mutate(
      contrast = contrast_name
    )

  Prok_DGE <- dplyr::bind_rows(
    Prok_DGE,
    contrast_result
  )
}

stopifnot(
  nrow(Prok_DGE) ==
    nrow(fit$coefficients) * ncol(contrasts)
)
```

```{r dge-annotation}
# -------------------------------------------------------------------------
# Annotate differential expression results
# -------------------------------------------------------------------------

Prok_DGE_annotated <- Prok_DGE %>%
  dplyr::left_join(
    Taxonomy,
    by = "orf"
  ) %>%
  dplyr::left_join(
    Prok_annotations,
    by = "orf"
  ) %>%
  dplyr::mutate(
    comparison = dplyr::case_when(
      stringr::str_starts(contrast, "R_vs_C") ~ "R vs C",
      stringr::str_starts(contrast, "RP_vs_C") ~ "R+P vs C",
      stringr::str_starts(contrast, "RP_vs_R") ~ "R+P vs R"
    ),
    time = dplyr::if_else(
      stringr::str_ends(contrast, "_0h"),
      "0h",
      "72h"
    )
  )

stopifnot(
  nrow(Prok_DGE_annotated) == nrow(Prok_DGE)
)

arrow::write_parquet(
  Prok_DGE_annotated,
    data_file("processed/prokaryotes/prok_differential_expression.parquet"))

```

```{r dge-overview, fig.width=10, fig.height=7}
# -------------------------------------------------------------------------
# Differential expression overview
# -------------------------------------------------------------------------
# This reproduces the logCPM-versus-logFC expression overview from the
# original analysis. It is not a classical volcano plot because the x-axis
# represents average expression rather than -log10(p).

Prok_DGE_plot <- Prok_DGE_annotated %>%
  dplyr::mutate(
    significant = FDR <= SIGNIFICANCE,
    transparency = dplyr::if_else(
      PValue < 0.05 & abs(logFC) > 1,
      1,
      0.1
    )
  )

Prok_DGE_overview_plot <- ggplot(
  Prok_DGE_plot,
  aes(
    x = logCPM,
    y = logFC,
    colour = FDR,
    shape = significant
  )
) +
  geom_point(
    aes(alpha = transparency),
    size = 0.8
  ) +
  scale_alpha_identity() +
  scale_colour_viridis_c("FDR") +
  facet_grid(
    time ~ comparison
  ) +
  geom_hline(
    yintercept = c(-1, 1),
    linetype = "dashed",
    linewidth = 0.3
  ) +
  theme_bw() +
  theme(
    legend.position = "top"
  ) +
  labs(
    x = "Average expression (logCPM)",
    y = "log2 fold change",
    caption = paste(
      "Differential expression overview.",
      "Highlighted ORFs have P < 0.05 and |logFC| > 1."
    )
  )

Prok_DGE_overview_plot

ggsave(
    "../results/prokaryotes/prok_differential_expression_overview.png",
  plot = Prok_DGE_overview_plot,
  width = 18,
  height = 12,
  units = "cm",
  dpi = 600
)
```
