---
title: "Eukaryotic Metatranscriptomic Analysis of Riverine and Phosphorus Additions"
author: "Erick Delgadillo"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    number_sections: true
---

# Overview

This document reproduces the eukaryotic metatranscriptomic analysis
used in the published study on microbial responses to riverine and
phosphorus additions.

The analysis focuses on the poly(A)-selected RNA fraction of the
control (C), river-water addition (R), and river-water plus phosphorus
addition (R+P) mesocosms at 0 h and 72 h after treatment. One R+P sample
at 0 h was excluded from the original analysis because of insufficient
sequencing depth, leaving two biological replicates for that group.

The workflow includes taxonomic and functional annotation, transcript
abundance processing, TPM-based analyses, and differential gene
expression analysis.

The original laboratory scripts contained data and analyses from two
independent experiments. This reproducibility workflow retains only the
samples and analyses corresponding to the experiment reported in the
associated publication.

```{r setup, include=FALSE}
# -------------------------------------------------------------------------
# Packages and analysis settings
# -------------------------------------------------------------------------

library(dplyr, warn.conflicts = FALSE)
library(ggplot2)

source(if (file.exists("../R/paths.R")) "../R/paths.R" else "R/paths.R")

SIGNIFICANCE <- 0.05

# processed_dir <- here::here("data", "processed", "eukaryotes")
# results_dir <- here::here("results", "eukaryotes")
# 
# dir.create(processed_dir, recursive = TRUE, showWarnings = FALSE)
# dir.create(results_dir, recursive = TRUE, showWarnings = FALSE)
```

# Environmental data

```{r environmental-data}
# -------------------------------------------------------------------------
# Environmental metadata
# -------------------------------------------------------------------------
# Bacterial production, primary production and nutrient concentrations are
# summarized for the treatment × time combinations used in the published
# experiment.

production_data <- read.csv(
    data_file("metadata/environmental/WP2_Spring_BP_PP_Mesocosmos.csv"),
  sep = ",",
  check.names = FALSE
) %>%
  tidyr::separate(
    sample,
    into = c("treatment", "time"),
    sep = "_"
  ) %>%
  dplyr::select(-errorstandar)

nutrients <- read.csv(
    data_file("metadata/environmental/WP2_Spring_Mesocosmos.csv"),
  sep = ";",
  check.names = FALSE
) %>%
  tidyr::separate(
    sample,
    into = c("treatment", "time"),
    sep = "_"
  ) %>%
  dplyr::filter(time %in% c("0h", "72h")) %>%
  dplyr::group_by(treatment, time, variable) %>%
  dplyr::summarise(
    mean = mean(value),
    .groups = "drop"
  )

Environmental_data <- dplyr::bind_rows(
  nutrients,
  production_data
) %>%
  dplyr::mutate(
    treatment = dplyr::recode(treatment, "RP" = "R+P")
  ) %>%
  tidyr::unite(
    sample,
    treatment,
    time,
    sep = "_",
    remove = TRUE
  ) %>%
  tidyr::pivot_wider(
    names_from = variable,
    values_from = mean
  )
```


# Taxonomic annotation

```{r taxonomy-load}
# -------------------------------------------------------------------------
# Load and clean eukaryotic taxonomic assignments
# -------------------------------------------------------------------------
# Placeholder and explicitly unclassified assignments are treated as missing,
# allowing downstream grouping to fall back to broader taxonomic ranks.

taxonomy <- readr::read_tsv(
  data_file("raw/eukaryotes/taxonomy_classification.tsv.gz"),
  show_col_types = FALSE
) %>%
  dplyr::rename(
    geneid = orf
  ) %>%
  dplyr::select(
    geneid,
    Domain,
    Phylum,
    Class,
    Order,
    Family,
    Genus,
    Species
  ) %>%
  dplyr::mutate(
    dplyr::across(
      c(Domain, Phylum, Class, Order, Family, Genus, Species),
      ~ ifelse(
          stringr::str_detect(
            .x,
            "_X|_[A-I]|uncl\\.|Unclassified"
          ),
          NA,
          .x
        )
    )
  )

species_of_interest <- c(
  "Alexandrium",
  "Scrippsiella",
  "Symbiodinium",
  "Thalassiosira",
  "Chaetoceros",
  "Skeletonema",
  "Pseudo-nitzschia",
  "Dinophysiales",
  "Micromonas",
  "Ostreococcus",
  "Mamiellales"
)

families_of_interest <- c(
  "Choreotrichida",
  "Tintinnida"
)

orders_of_interest <- c(
  "Spirotrichea"
)
```

```{r taxonomy-consensus}
# -------------------------------------------------------------------------
# Hierarchical consensus for duplicated gene assignments
# -------------------------------------------------------------------------
# Some genes have more than one taxonomic assignment. Missing information in
# one assignment is not considered a conflict. A conflict occurs only when
# different non-missing values are assigned to the same rank.

Taxonomy_consensus <- taxonomy %>%
  dplyr::group_by(geneid) %>%
  dplyr::summarise(
    conflict_Domain = dplyr::n_distinct(Domain, na.rm = TRUE) > 1,
    Domain = dplyr::first(
      Domain[!is.na(Domain)],
      default = NA_character_
    ),

    conflict_Phylum = dplyr::n_distinct(Phylum, na.rm = TRUE) > 1,
    Phylum = dplyr::first(
      Phylum[!is.na(Phylum)],
      default = NA_character_
    ),

    conflict_Class = dplyr::n_distinct(Class, na.rm = TRUE) > 1,
    Class = dplyr::first(
      Class[!is.na(Class)],
      default = NA_character_
    ),

    conflict_Order = dplyr::n_distinct(Order, na.rm = TRUE) > 1,
    Order = dplyr::first(
      Order[!is.na(Order)],
      default = NA_character_
    ),

    conflict_Family = dplyr::n_distinct(Family, na.rm = TRUE) > 1,
    Family = dplyr::first(
      Family[!is.na(Family)],
      default = NA_character_
    ),

    conflict_Genus = dplyr::n_distinct(Genus, na.rm = TRUE) > 1,
    Genus = dplyr::first(
      Genus[!is.na(Genus)],
      default = NA_character_
    ),

    conflict_Species = dplyr::n_distinct(Species, na.rm = TRUE) > 1,
    Species = dplyr::first(
      Species[!is.na(Species)],
      default = NA_character_
    ),

    .groups = "drop"
  ) %>%
  dplyr::mutate(
    Domain = dplyr::if_else(
      conflict_Domain,
      NA_character_,
      Domain
    ),

    Phylum = dplyr::if_else(
      conflict_Domain | conflict_Phylum,
      NA_character_,
      Phylum
    ),

    Class = dplyr::if_else(
      conflict_Domain | conflict_Phylum | conflict_Class,
      NA_character_,
      Class
    ),

    Order = dplyr::if_else(
      conflict_Domain | conflict_Phylum | conflict_Class |
        conflict_Order,
      NA_character_,
      Order
    ),

    Family = dplyr::if_else(
      conflict_Domain | conflict_Phylum | conflict_Class |
        conflict_Order | conflict_Family,
      NA_character_,
      Family
    ),

    Genus = dplyr::if_else(
      conflict_Domain | conflict_Phylum | conflict_Class |
        conflict_Order | conflict_Family | conflict_Genus,
      NA_character_,
      Genus
    ),

    Species = dplyr::if_else(
      conflict_Domain | conflict_Phylum | conflict_Class |
        conflict_Order | conflict_Family | conflict_Genus |
        conflict_Species,
      NA_character_,
      Species
    )
  )
```

```{r taxonomy-final}
# -------------------------------------------------------------------------
# Final taxonomic grouping
# -------------------------------------------------------------------------
# Taxonomic resolution is intentionally heterogeneous, preserving the groups
# used in the original eukaryotic analysis.

Taxonomy <- Taxonomy_consensus %>%
  dplyr::mutate(
    Final_Taxonomy = dplyr::case_when(
      Phylum == "Amoebozoa" ~ paste0(Phylum),
      Phylum == "Excavata" ~ paste0(Domain),
      Phylum == "Hacrobia" ~ paste0(Phylum),
      Phylum == "Opisthokonta" ~ paste0(Phylum),
      Phylum == "Rhizaria" ~ paste0(Phylum),

      Family == "Suessiales" &
        Species %in% species_of_interest ~ paste0(Species),
      Family == "Suessiales" &
        !Species %in% species_of_interest ~ paste0(Family),

      Family == "Peridiniales" &
        Species %in% species_of_interest ~ paste0(Species),
      Family == "Peridiniales" &
        !Species %in% species_of_interest ~ paste0(Family),

      Family == "Gymnodiniales" &
        Species %in% species_of_interest ~ paste0(Species),
      Family == "Gymnodiniales" &
        !Species %in% species_of_interest ~ paste0(Family),

      Family == "Gonyaulacales" &
        Species %in% species_of_interest ~ paste0(Species),
      Family == "Gonyaulacales" &
        !Species %in% species_of_interest ~ paste0(Family),

      Family == "Prorocentrales" &
        Species %in% species_of_interest ~ paste0(Species),
      Family == "Prorocentrales" &
        !Species %in% species_of_interest ~ paste0(Family),

      Class == "Dinoflagellata" &
        !Species %in% species_of_interest ~ paste0(Order),

      Order == "Dinophyceae" &
        !Species %in% species_of_interest ~ paste0(Family),

      Family == "Choreotrichida" ~ paste0(Family),
      Family == "Tintinnida" ~ paste0(Family),

      Order == "Spirotrichea" &
        !Family %in% families_of_interest ~ paste0(Order),

      Class == "Ciliophora" &
        !Order %in% orders_of_interest ~ paste0(Class),

      Order == "Bacillariophyta" &
        Species %in% species_of_interest ~ paste0(Species),

      Order == "Bacillariophyta" &
        !Species %in% species_of_interest ~ paste0(Order),

      Phylum == "Stramenopiles" &
        Order != "Bacillariophyta" ~ paste0(Class),

      Class == "Ochrophyta" &
        is.na(Order) ~ paste0(Class),

      Phylum == "Stramenopiles" &
        is.na(Class) ~ paste0(Phylum),

      Phylum == "Archaeplastida" &
        Species %in% species_of_interest ~ paste0(Species),

      Phylum == "Archaeplastida" &
        !Species %in% species_of_interest ~ paste0(Class)
    )
  ) %>%
  dplyr::mutate(
    Final_Taxonomy = dplyr::case_when(
      is.na(Final_Taxonomy) ~ paste0(Phylum),
      !is.na(Final_Taxonomy) ~ Final_Taxonomy
    ),
    Final_Taxonomy = ifelse(
      Final_Taxonomy == "NA",
      paste0(Class),
      Final_Taxonomy
    ),
    Final_Taxonomy = ifelse(
      Final_Taxonomy == "NA",
      paste0(Phylum),
      Final_Taxonomy
    ),
    Final_Taxonomy = ifelse(
      Final_Taxonomy == "NA",
      paste0(Domain),
      Final_Taxonomy
    ),
    Final_Taxonomy = dplyr::case_when(
      Final_Taxonomy == "Prorocentrales" ~ "Dinophyceae",
      Final_Taxonomy == "Pelagodinium" ~ "Suessiaceae",
      Final_Taxonomy == "Noctilucophyceae" ~ "Dinoflagellata",
      Final_Taxonomy == "Oxyrrhea" ~ "Dinoflagellata",
      Final_Taxonomy == "Opalozoa" ~ "Stramenopiles",
      Final_Taxonomy == "Archaeplastida" ~ "Eukaryota",
      Final_Taxonomy == "Alveolata" ~ "Eukaryota",
      Final_Taxonomy == "Ciliopora" ~ "Eukaryota",
      TRUE ~ Final_Taxonomy
    )
  ) %>%
  dplyr::select(
    geneid,
    Domain,
    Phylum,
    Class,
    Order,
    Family,
    Genus,
    Species,
    Final_Taxonomy
  )

stopifnot(!anyDuplicated(Taxonomy$geneid))
```

# Sample metadata

```{r sample-metadata}
# -------------------------------------------------------------------------
# Eukaryotic sample metadata
# -------------------------------------------------------------------------
# Sequencing identifiers are linked to the experimental design. Only samples
# belonging to the experiment reproduced in this repository are retained.

samples_tags <- readr::read_csv(
  data_file("metadata/eukaryotes/INTERES_Euk_samples_tags.csv"),
  show_col_types = FALSE
) %>%
  dplyr::select(
    sample,
    Sample,
    group,
    date
  ) %>%
  tidyr::separate(
    Sample,
    into = c(
      "Workpackage",
      "experiment",
      "treatment",
      "time",
      "replic"
    ),
    sep = "_",
    remove = FALSE
  ) %>%
  dplyr::mutate(
    sample = as.character(sample)
  ) %>%
  dplyr::filter(Workpackage == "WP2")

replicate_counts <- samples_tags %>%
  dplyr::count(
    treatment,
    time,
    name = "n_replicates"
  )

stopifnot(
  nrow(samples_tags) == 17,
  !anyDuplicated(samples_tags$sample)
)
```

# Functional annotation

```{r functional-annotation}
# -------------------------------------------------------------------------
# eggNOG functional annotation
# -------------------------------------------------------------------------
# Only the identifiers required downstream are retained. Up to three KEGG
# ortholog assignments are preserved in priority order.

Euk_annotations <- suppressWarnings(
  data.table::fread(
      data_file("raw/eukaryotes/all_samples_emapper_annotations.tsv"))
) %>%
  dplyr::rename(
    geneid = `#query`
  ) %>%
  dplyr::select(
    geneid,
    KEGG_ko,
    KEGG_Module,
    PFAMs
  ) %>%
  tidyr::separate(
    KEGG_ko,
    into = c("KEGG_ko", "KEGG_ko2", "KEGG_ko3"),
    sep = ",",
    remove = TRUE,
    extra = "drop",
    fill = "right"
  ) %>%
  dplyr::mutate(
    KEGG_ko = stringr::str_remove(KEGG_ko, "ko:"),
    KEGG_ko2 = stringr::str_remove(KEGG_ko2, "ko:"),
    KEGG_ko3 = stringr::str_remove(KEGG_ko3, "ko:"),
    KEGG_Module = sub(",.*$", "", KEGG_Module)
  )

stopifnot(!anyDuplicated(Euk_annotations$geneid))
```

# Transcript abundance

```{r counts-and-tpm}
# -------------------------------------------------------------------------
# Raw counts, eukaryotic filtering and TPM recalculation
# -------------------------------------------------------------------------
# Counts are restricted to the 17 study samples and to genes represented in
# the curated eukaryotic taxonomy. TPM is recalculated within this retained
# eukaryotic fraction.

euk_counts_raw <- readr::read_tsv(
  data_file("raw/eukaryotes/polyA_counts.tsv.gz"),
  show_col_types = FALSE
) %>%
  dplyr::rename(
    geneid = Geneid
  ) %>%
  dplyr::mutate(
    geneid = stringr::str_remove(geneid, "^cds\\."),
    sample = as.character(sample)
  ) %>%
  dplyr::filter(
    sample %in% samples_tags$sample
  ) %>%
  dplyr::semi_join(
    Taxonomy %>%
      dplyr::filter(Domain == "Eukaryota"),
    by = "geneid"
  ) %>%
  dplyr::group_by(sample) %>%
  dplyr::mutate(
    tpm = (count / sum(count)) * 1e6
  ) %>%
  dplyr::ungroup()

stopifnot(
  setequal(unique(euk_counts_raw$sample), samples_tags$sample)
)

tpm_check <- euk_counts_raw %>%
  dplyr::group_by(sample) %>%
  dplyr::summarise(
    total_tpm = sum(tpm),
    .groups = "drop"
  )

stopifnot(
  all(abs(tpm_check$total_tpm - 1e6) < 1)
)
```

# Protein profiles of interest

```{r protein-profiles}
# -------------------------------------------------------------------------
# Functional protein profiles
# -------------------------------------------------------------------------
# Descriptive names not used downstream are discarded before duplicate
# removal. This prevents duplicated TPM values when two profile rows differ
# only in the unused `name` field.

Hiper_proteins_analysis <- readxl::read_excel(
    data_file("metadata/environmental/Protein_profiles_2024.xlsx"),
  sheet = "Protein profiles",
  col_names = TRUE
) %>%
  dplyr::select(
    Category,
    Subcategory,
    SubcategoryP,
    gene_protein,
    KEGG_ko,
    kingdom
  ) %>%
  dplyr::filter(!is.na(KEGG_ko)) %>%
  dplyr::distinct()

functional_kos <- unique(Hiper_proteins_analysis$KEGG_ko)
```

# Integration of expression and annotations

```{r annotation-joins}
# -------------------------------------------------------------------------
# Join expression, taxonomy, functional annotation and sample metadata
# -------------------------------------------------------------------------

Euk_annotated <- euk_counts_raw %>%
  dplyr::left_join(
    Taxonomy,
    by = "geneid"
  ) %>%
  dplyr::left_join(
    Euk_annotations,
    by = "geneid"
  ) %>%
  dplyr::left_join(
    samples_tags,
    by = "sample"
  )

stopifnot(
  nrow(Euk_annotated) == nrow(euk_counts_raw)
)
```

```{r annotated-counts-tpm}
# -------------------------------------------------------------------------
# Tables used by downstream analyses
# -------------------------------------------------------------------------
# There is one observation per gene and biological replicate at this stage,
# so the original group_by()/summarise() operations reduce to a direct
# projection of the required columns.

Euk_counts_annotated <- Euk_annotated %>%
  dplyr::transmute(
    Workpackage,
    treatment,
    time,
    replic,
    geneid,
    Final_Taxonomy,
    Phylum,
    Class,
    Order,
    Family,
    Genus,
    Species,
    KEGG_ko,
    KEGG_ko2,
    KEGG_ko3,
    KEGG_Module,
    PFAMs,
    count,
    nReplics = 1L
  )

Euk_TPMs_annotated <- Euk_annotated %>%
  dplyr::transmute(
    Workpackage,
    treatment,
    time,
    replic,
    geneid,
    Final_Taxonomy,
    Domain,
    Phylum,
    Class,
    Order,
    Family,
    Genus,
    Species,
    KEGG_ko,
    KEGG_ko2,
    KEGG_ko3,
    KEGG_Module,
    PFAMs,
    tpm,
    nReplics = 1L
  )

arrow::write_parquet(
  Euk_counts_annotated,
    data_file("processed/eukaryotes/euk_counts_annotated.parquet"))

arrow::write_parquet(
    Euk_TPMs_annotated,
    data_file("processed/eukaryotes/euk_tpms_annotated.parquet"))
```

# Functional TPM analysis

```{r functional-selection}
# -------------------------------------------------------------------------
# Select genes of functional interest using prioritized KEGG assignments
# -------------------------------------------------------------------------
# For each gene, the first matching KEGG assignment is retained in the order
# KEGG_ko -> KEGG_ko2 -> KEGG_ko3.

Euk_filtered <- Euk_TPMs_annotated %>%
  dplyr::mutate(
    selected_KEGG = dplyr::case_when(
      KEGG_ko %in% functional_kos ~ KEGG_ko,
      KEGG_ko2 %in% functional_kos ~ KEGG_ko2,
      KEGG_ko3 %in% functional_kos ~ KEGG_ko3,
      TRUE ~ NA_character_
    )
  ) %>%
  dplyr::filter(!is.na(selected_KEGG)) %>%
  dplyr::mutate(
    KEGG_ko = selected_KEGG
  ) %>%
  dplyr::select(-selected_KEGG)

Euk_functional_joined <- Euk_filtered %>%
  dplyr::left_join(
    Hiper_proteins_analysis,
    by = "KEGG_ko",
    relationship = "many-to-many"
  ) %>%
  dplyr::filter(
    !kingdom %in% c(
      "Bacteria",
      "Bacteria/Archaea",
      "Archaea"
    )
  )
```

```{r functional-tpm}
# -------------------------------------------------------------------------
# Aggregate TPM by functional gene and taxonomic group
# -------------------------------------------------------------------------

Euk_functional_TPM <- Euk_functional_joined %>%
  dplyr::group_by(
    Workpackage,
    treatment,
    time,
    replic,
    gene_protein,
    KEGG_ko,
    Category,
    Subcategory,
    SubcategoryP,
    Final_Taxonomy
  ) %>%
  dplyr::summarise(
    tpm = sum(tpm),
    .groups = "drop"
  )

arrow::write_parquet(
  Euk_functional_TPM,
    data_file("processed/eukaryotes/euk_functional_tpms.parquet"))
```

# Mean TPM and taxonomic composition

```{r mean-tpm}
# -------------------------------------------------------------------------
# Mean TPM across biological replicates
# -------------------------------------------------------------------------
# Most treatment x time groups have three biological replicates. R+P at 0 h
# has two, so the divisor is taken directly from the sample metadata.

Euk_mean_TPM <- Euk_TPMs_annotated %>%
  dplyr::left_join(
    replicate_counts,
    by = c("treatment", "time")
  ) %>%
  dplyr::group_by(
    Workpackage,
    treatment,
    time,
    n_replicates,
    geneid,
    Final_Taxonomy,
    Phylum,
    Class,
    Order,
    Family,
    Genus,
    Species,
    KEGG_ko,
    KEGG_ko2,
    KEGG_ko3,
    KEGG_Module,
    PFAMs
  ) %>%
  dplyr::summarise(
    tpm = sum(tpm) / dplyr::first(n_replicates),
    nReplics_detected = dplyr::n(),
    .groups = "drop"
  ) %>%
  dplyr::select(-n_replicates)

arrow::write_parquet(
  Euk_mean_TPM,
    data_file("processed/eukaryotes/euk_mean_tpms.parquet"))
```

```{r taxonomic-composition}
# -------------------------------------------------------------------------
# Taxonomic composition of the eukaryotic metatranscriptome
# -------------------------------------------------------------------------

Euk_MetaT_Proportion_FinalTax <- Euk_mean_TPM %>%
  tidyr::unite(
    sample,
    treatment,
    time,
    sep = "_",
    remove = FALSE
  ) %>%
  dplyr::group_by(
    sample,
    Final_Taxonomy
  ) %>%
  dplyr::summarise(
    Total_tpm = sum(tpm),
    .groups = "drop"
  )
```

# Functional gene ratios

```{r gene-ratios}
# -------------------------------------------------------------------------
# Functional gene ratios within taxonomic groups
# -------------------------------------------------------------------------
# Functional TPM values are averaged using the number of biological
# replicates actually available for each treatment x time group.
#
# pstA, pstB, pstC and pstS are grouped as components of the pstABCS
# high-affinity phosphate transport system.

Euk_gene_ratios <- Euk_functional_TPM %>%
  dplyr::mutate(
    gene_protein = dplyr::if_else(
      gene_protein %in% c("pstA", "pstB", "pstC", "pstS"),
      "pstABCS",
      gene_protein
    )
  ) %>%
  dplyr::left_join(
    replicate_counts,
    by = c("treatment", "time")
  ) %>%
  dplyr::group_by(
    treatment,
    time,
    n_replicates,
    gene_protein,
    Category,
    Subcategory,
    SubcategoryP,
    Final_Taxonomy
  ) %>%
  dplyr::summarise(
    tpm = sum(tpm) / dplyr::first(n_replicates),
    .groups = "drop"
  ) %>%
  dplyr::select(-n_replicates) %>%
  tidyr::unite(
    sample,
    treatment,
    time,
    sep = "_",
    remove = TRUE
  ) %>%
  dplyr::left_join(
    Euk_MetaT_Proportion_FinalTax,
    by = c("sample", "Final_Taxonomy")
  ) %>%
  dplyr::mutate(
    GenRatio = tpm / Total_tpm
  )

stopifnot(
  !any(is.na(Euk_gene_ratios$Total_tpm)),
  !any(is.na(Euk_gene_ratios$GenRatio)),
  !any(is.infinite(Euk_gene_ratios$GenRatio))
)

readr::write_csv(
  Euk_gene_ratios,
    data_file("processed/eukaryotes/euk_gene_ratios.csv"))

```

# Differential gene expression analysis

Differential gene expression was evaluated from raw eukaryotic read
counts using edgeR. Only samples belonging to the experiment described
in the associated publication are included in this analysis.

```{r dge-count-matrix}
# -------------------------------------------------------------------------
# Count matrix and experimental groups
# -------------------------------------------------------------------------
# Genes not detected in a sample are assigned a count of zero. Sample columns
# are explicitly ordered according to the metadata.

Euk_count_matrix <- euk_counts_raw %>%
  dplyr::select(
    geneid,
    sample,
    count
  ) %>%
  tidyr::pivot_wider(
    names_from = sample,
    values_from = count,
    values_fill = 0
  ) %>%
  tibble::column_to_rownames("geneid")

Euk_count_matrix <- Euk_count_matrix[
  ,
  samples_tags$sample,
  drop = FALSE
]

dge_samples <- samples_tags %>%
  dplyr::mutate(
    group = factor(
      paste(treatment, time, sep = "_"),
      levels = c(
        "C_0h",
        "R_0h",
        "R+P_0h",
        "C_72h",
        "R_72h",
        "R+P_72h"
      )
    )
  )

stopifnot(
  identical(
    colnames(Euk_count_matrix),
    dge_samples$sample
  )
)
```

```{r dge-model}
# -------------------------------------------------------------------------
# edgeR filtering, normalization and model fitting
# -------------------------------------------------------------------------

dge_samples_df <- as.data.frame(dge_samples)
rownames(dge_samples_df) <- dge_samples_df$sample

dgelist <- edgeR::DGEList(
  counts = Euk_count_matrix,
  samples = dge_samples_df
)

keep <- edgeR::filterByExpr(
  dgelist,
  group = dgelist$samples$group
)

dgelist <- dgelist[
  keep,
  ,
  keep.lib.sizes = FALSE
]

dgelist <- edgeR::calcNormFactors(
  dgelist,
  method = "TMM"
)

design <- model.matrix(
  ~ 0 + group,
  data = dgelist$samples
)

colnames(design) <- c(
  "C_0h",
  "R_0h",
  "RP_0h",
  "C_72h",
  "R_72h",
  "RP_72h"
)

stopifnot(
  identical(
    rownames(design),
    rownames(dgelist$samples)
  )
)

cdisp <- edgeR::estimateDisp(
  dgelist,
  design
)

fit <- edgeR::glmQLFit(
  cdisp,
  design
)
```

```{r dge-contrasts}
# -------------------------------------------------------------------------
# Differential expression contrasts
# -------------------------------------------------------------------------
# Positive logFC values indicate higher expression in the first group named
# in each contrast.

contrasts <- limma::makeContrasts(
  R_vs_C_0h = R_0h - C_0h,
  RP_vs_C_0h = RP_0h - C_0h,
  RP_vs_R_0h = RP_0h - R_0h,
  R_vs_C_72h = R_72h - C_72h,
  RP_vs_C_72h = RP_72h - C_72h,
  RP_vs_R_72h = RP_72h - R_72h,
  levels = design
)
```

```{r dge-tests}
# -------------------------------------------------------------------------
# Quasi-likelihood tests and FDR correction
# -------------------------------------------------------------------------
# The six explicitly defined contrasts are tested in the order shown above.

Euk_DGE <- tibble::tibble()

for (contrast_name in colnames(contrasts)) {

  test <- edgeR::glmQLFTest(
    fit,
    contrast = contrasts[, contrast_name]
  )

  contrast_result <- edgeR::topTags(
    test,
    n = Inf,
    adjust.method = "fdr"
  )$table %>%
    tibble::rownames_to_column("geneid") %>%
    dplyr::mutate(
      contrast = contrast_name
    )

  Euk_DGE <- dplyr::bind_rows(
    Euk_DGE,
    contrast_result
  )
}

stopifnot(
  nrow(Euk_DGE) ==
    nrow(fit$coefficients) * ncol(contrasts)
)
```

```{r dge-annotation}
# -------------------------------------------------------------------------
# Annotate differential expression results
# -------------------------------------------------------------------------

Euk_DGE_annotated <- Euk_DGE %>%
  dplyr::left_join(
    Taxonomy,
    by = "geneid"
  ) %>%
  dplyr::left_join(
    Euk_annotations,
    by = "geneid"
  ) %>%
  dplyr::mutate(
    comparison = dplyr::case_when(
      stringr::str_starts(contrast, "R_vs_C") ~ "R vs C",
      stringr::str_starts(contrast, "RP_vs_C") ~ "R+P vs C",
      stringr::str_starts(contrast, "RP_vs_R") ~ "R+P vs R"
    ),
    time = dplyr::if_else(
      stringr::str_ends(contrast, "_0h"),
      "0h",
      "72h"
    )
  )

stopifnot(
  nrow(Euk_DGE_annotated) == nrow(Euk_DGE)
)

arrow::write_parquet(
  Euk_DGE_annotated,
    data_file("processed/eukaryotes/euk_differential_expression.parquet"))

```

```{r dge-overview, fig.width=10, fig.height=7}
# -------------------------------------------------------------------------
# Differential expression overview
# -------------------------------------------------------------------------
# This reproduces the logCPM-versus-logFC expression overview from the
# original analysis. It is not a classical volcano plot because the x-axis
# represents average expression rather than -log10(p).

Euk_DGE_plot <- Euk_DGE_annotated %>%
  dplyr::mutate(
    significant = FDR <= SIGNIFICANCE,
    transparency = dplyr::if_else(
      PValue < 0.05 & abs(logFC) > 1,
      1,
      0.1
    )
  )

Euk_DGE_overview_plot <- ggplot(
  Euk_DGE_plot,
  aes(
    x = logCPM,
    y = logFC,
    colour = FDR,
    shape = significant
  )
) +
  geom_point(
    aes(alpha = transparency),
    size = 0.8
  ) +
  scale_alpha_identity() +
  scale_colour_viridis_c("FDR") +
  facet_grid(
    time ~ comparison
  ) +
  geom_hline(
    yintercept = c(-1, 1),
    linetype = "dashed",
    linewidth = 0.3
  ) +
  theme_bw() +
  theme(
    legend.position = "top"
  ) +
  labs(
    x = "Average expression (logCPM)",
    y = "log2 fold change",
    caption = paste(
      "Differential expression overview.",
      "Highlighted genes have P < 0.05 and |logFC| > 1."
    )
  )

Euk_DGE_overview_plot

ggsave(
    "../results/eukaryotes/euk_differential_expression_overview.png",
  plot = Euk_DGE_overview_plot,
  width = 18,
  height = 12,
  units = "cm",
  dpi = 600
)
```
